---
title: "ECON 773: Assignment 4"
author: "The BLUE Team, Jeneta Ljutic (400138620), Tadhg Taylor-McGreal (400330297), Stella Till (400364649)"
date: "February 24, 2026"
format:
  typst: 
    toc: TRUE
---

{{< pagebreak >}} 

# Preface

## Goal

The goals of this assignment are to:

- use difference-in-differences methods to analyze treatments with unobserved confounders
- practice DiD mechanics
- review IV mechanics and interpretation

## Instructions

See assignment 1.

We will use the same tools as in Assignments 1--3:

- `tidyverse` for data manipulation and plotting
- `haven` to load Stata data files
- `gt` and `gtsummary` for tables
- `estimatr` for regression (`lm_robust`)

```{r}
#| include: false
library(tidyverse)
library(haven)
library(gt)
library(gtsummary)
library(estimatr)
```

{{< pagebreak >}} 

# Card and Krueger (1994)

You will replicate [Card and Krueger, AER, 1994](https://davidcard.berkeley.edu/papers/njmin-aer.pdf). The data are in `CK1994.dta`, which is documented in `CK1994_description.pdf`.

Some context:

- On April 1, 1992, the minimum wage in New Jersey rose from USD 4.25 to USD 5.05, while Pennsylvania’s minimum wage stayed at USD 4.25.
- The original survey was conducted in February 1992 (pre) and November 1992 (post), covering 331 restaurants in NJ and 79 in PA.
- We use a long-format panel with two rows per restaurant and therefore cluster by `store`.

```{r}
ck_df <- read_dta("CK1994.dta") |>
  as_tibble() |>
  mutate(
    nj = state == 1,
    post = time == 1,
    emp_fte = empft + 0.5 * emppt + nmgrs,
    pmeal = meals
  ) |> 
  mutate(
    state = if_else(nj, "New Jersey", "Pennsylvania"),
    chain = as_factor(chain)
  )
```

Note: 

- `mutate(...)` creates a few convenience variables we’ll use throughout:
  - `nj`: `TRUE` for New Jersey restaurants (`state == 1`), `FALSE` for Pennsylvania.
  - `post`: `TRUE` for the post-policy survey wave (`time == 1`), `FALSE` for pre (`time == 0`).
  - `emp_fte`: a "full-time equivalent" employment measure: full-time workers (`empft`) + half of part-time workers (`0.5 * emppt`) + managers (`nmgrs`).
  - `pmeal`: the meal price variable

## Restaurant chains

For each state, calculate the percentage of Burger King, KFC, Roys, and Wendys in the sample. Is the distribution of restaurant chains comparable across the two states?

### Answer

```{r}
library(dplyr)

ck_df |> 
  mutate(
    chain = factor(chain,
                   levels = 1:4,
                   labels = c("Burger King", "KFC", "Roys", "Wendys")),
    state = factor(state)
  ) |> 
  group_by(state, chain) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(state) |> 
  mutate(percent = 100 * n / sum(n)) |> 
  arrange(state, desc(percent))
```

In this set of code, we first load dplyr. Then we use the dataset ck_df. We then use "mutate", to convert the data into labeled categories and the state variable. The command "chain", groups the data by state and chain to count observations. Then, we direct the code to once again "group_by" state and compute each chain's percentage share within that state using "mutate". Lastly, "arrage" sorts the results to appear from the highest to lowest share in each state.

These results show the distribution of restaurant chains in the sample for New Jersey and Pennsylvania, expressed as percentages within each state. In both states, Burger King makes up the largest share (about 41% in New Jersey and 44% in Pennsylvania), while Wendy’s has the smallest share in New Jersey but a somewhat larger share in Pennsylvania. Roy’s and KFC account for moderate portions of the sample, with Roy’s slightly more common in New Jersey and KFC less common in Pennsylvania. Overall, the distributions are fairly similar across the two states, suggesting the samples are reasonably comparable in terms of restaurant chain composition.

```{r}
ck_df |> group_by(state) |> summarize(n_obs = n())
```
The number of observations in New Jersey is 662. The number of observations in Pennsylvania is 158. 

## Wage change

Compute the average wage in each state, before and after the minimum wage increase. What does this suggest about the adoption of the minimum-wage policy in NJ?

### Answer

```{r}
ck_df |> group_by(state, time) |> 
  summarize(
    m_wage = mean(wage_st, na.rm = TRUE)
  )
```

In the code block we first take the dataset "ck_df" and pass it down to the next operation using the pipe operator. Then, we group the data by state and time. This means that all calculations are done separately for each state-time combination (i.e. NJ before, NJ after, PA before, PA after). Then, using "summarize(m_wage = mean(wage_st, na.rm = TRUE))", we collapse each group into a single row by computing the average of the variable wage_st within each group. na.rm = TRUE tells R to ignore missing values when calculating the mean.

NJ minium wage increased from \$4.61 to \$5.08. The PA minium wage (the control group), stayed relatively the same increases from \$4.63 to \$4.62. This sharp increase in New Jersey but not in Pennsylvania is consistent with the implementation of the minimum-wage increase in NJ only. Therefore, the results suggest that the policy was successfully adopted and had a clear effect on wages in New Jersey.

## DiD for wages

Calculate the DiD estimate for wages manually. Then compute the same estimate using `lm_robust`. Interpret the result.

### Answer

```{r}
lm_robust(
  wage_st ~ nj + post + nj*post, data = ck_df
)
```

```{r}
lm_robust(
  emp_fte ~ nj*post, data = ck_df
)
```

Both lines estimate Difference-in-Differences regressions using lm_robust(), which fits linear models with robust standard errors on the dataset ck_df. The first model examines how the policy affected starting wages (wage_st), while the second examines its effect on full-time-equivalent employment (emp_fte), using New Jersey, the post-policy period, and their interaction. We use lm_robust() since it fits a linear regression with robust standard errors (from the estimatr package), which makes inference less sensitive to heteroskedasticity.

The minimum-wage increase raised average wages in New Jersey by about $0.48 more than in Pennsylvania. This effect is large and highly statistically significant. Therefore, the results indicate that the policy substantially increased wages in New Jersey relative to the control state.

## Employment effect

Repeat the analysis for employment. Use `emp_fte` (full-time equivalent) as the outcome. Interpret the DiD estimate.

### Answer

```{r}
lm_robust(
  wage_st ~ nj + post + nj*post, data = ck_df
)
```

```{r}
lm_robust(
  emp_fte ~ nj*post, data = ck_df
)
```

The Difference-in-Differences estimate for employment is the coefficient on njTRUE:postTRUE, which is approximately 2.75. This suggests that, after the minimum wage increase, employment in New Jersey restaurants rose by about 2.75 full-time equivalent workers per restaurant relative to Pennsylvania. However, this estimate is not statistically significant (p ≈ 0.13), and the confidence interval includes zero. Therefore, there is no strong evidence that the minimum wage increase had a meaningful effect on employment.

## Interpreting DiD

Under what conditions does this estimate identify the ATT? Do we have any suggestive evidence to support or refute these conditions? Why can we not run a pre-trend test in this application? What would we need to test pre-trends?

### Answer

The two-period DiD estimate identifies the ATT under the following assumptions. Firstly, parallel trends must hold. This means that absent the minimum wage increase, employment in New Jersey would have followed the same time trend as employment in Pennsylvania  and if there are no spillovers and no differential changes in sample composition across states over time (SUTVA). In this application we can provide only suggestive evidence: NJ and PA have similar pre-policy wages and are in the same industry and region, and wages increase sharply in NJ post-policy but not in PA, indicating the policy was implemented. We cannot test the parallel-trends assumption directly because there is only one pre-treatment period, so there is no pre-trend to estimate. Testing pre-trends would require employment outcomes in multiple pre-policy periods, enabling an event-study (lead) test or comparisons of pre-period changes between NJ and PA.

The DiD estimate also relies on a no-anticipation assumption: restaurants in New Jersey must not have adjusted employment prior to the policy in response to the expected minimum wage increase. If firms reduced staffing before April 1992, the pre-treatment period would already reflect treatment effects, biasing the estimate. In this study, the pre survey occurred in February 1992, only about two months before implementation, suggesting limited time for adjustment. However, because there is only one pre-treatment observation, we cannot test for anticipatory behavior directly. Doing so would require employment data from earlier pre-policy periods to determine whether NJ outcomes began diverging from PA before the law took effect.

## Meal prices

Compute the DiD estimator for the average meal price (`pmeal`). Do restaurants subject to the wage increase raise prices?

### Answer

```{r}
lm_robust(
  wage_st ~ nj + post + nj*post, data = ck_df
)
```

```{r}
lm_robust(
  pmeal ~ nj*post, data = ck_df
)
```

The DiD estimator for the average meal price (pmeal) is approximately −0.0195. This indicates that meal prices in New Jersey restaurants decreased slightly relative to those in Pennsylvania after the minimum wage increase. This contradicts standard economic theory that would suggest for prices to rise after a wage change. However, since the estimate is very small and not statistically significant, as the confidence interval includes zero we can conclude that there is no evidence that restaurants subject to the wage increase raised prices.

## Another outcome

Open `CK1994_description.pdf` and choose another outcome variable (other than wages, employment, or meal prices). Describe the variable and run a DiD analysis for it. Interpret your result.

### Answer

We first chose the variable "open", to test if the hour which a restaurant opens changes after the minimum wage increase in NJ. We follow the same steps as in the sections above. 

```{r}
lm_robust(
  wage_st ~ nj + post + nj*post, data = ck_df
)
```

```{r}
lm_robust(
  open ~ nj*post, data = ck_df
)
```

The Difference-in-Differences estimate for restaurant opening time is approximately 0.015 hours, indicating that New Jersey restaurants opened about one minute later relative to Pennsylvania after the minimum wage increase. However, this effect is extremely small and statistically insignificant (p ≈ 0.97), with a confidence interval that includes zero. Therefore, there is no evidence that the policy affected opening hours. This suggests that restaurants did not adjust operating schedules in response to the wage increase.

Next, we test if the number of part time employees changed as a result of the minimum wage change in NJ.

```{r}
lm_robust(
  wage_st ~ nj + post + nj*post, data = ck_df
)
```

```{r}
lm_robust(
  emppt ~ nj*post, data = ck_df
)
```

The Difference-in-Differences estimate for part-time employment is −0.77, indicating that New Jersey restaurants employed slightly fewer part-time workers relative to Pennsylvania after the minimum wage increase. However, this effect is not statistically significant (p ≈ 0.67), and the confidence interval includes zero. Therefore, there is no evidence that the policy reduced part-time employment. Overall, the results suggest that the minimum wage increase did not meaningfully affect part-time staffing levels.

Lastly, we test if the number of cash registers in the store changed as a result of the minium wage increase. This channel tests if the restaurants moved employees to the back of the store for production and reduced the number of employees needed to fill the spot at a cashregister. 

```{r}
lm_robust(
  wage_st ~ nj + post + nj*post, data = ck_df
)
```

```{r}
lm_robust(
 nregisters  ~ nj*post, data = ck_df
)
```

The Difference-in-Differences estimate for the number of registers is −0.13, indicating that New Jersey restaurants had slightly fewer registers relative to Pennsylvania after the minimum wage increase. However, this effect is not statistically significant (p ≈ 0.53), and the confidence interval includes zero. Therefore, there is no evidence that the policy affected the number of registers. This suggests that restaurants did not adjust their capital equipment in response to the wage increase.

{{< pagebreak >}} 

# Dinas et al. (2018)

You will replicate some findings from [Dinas et al. (2018)](https://www.cambridge.org/core/journals/political-analysis/article/waking-up-the-golden-dawn-does-exposure-to-the-refugee-crisis-increase-support-for-extremeright-parties/C50A127CC517968F2D0FA42A2A23FF85).
The data are in `dinas_golden_dawn.Rdata`, via [Harvard dataverse](https://doi.org/10.7910/DVN/3IWTGB).

Context (via [de Rosemont](https://uclspp.github.io/PUBL0050/5-panel-data-and-difference-in-differences.html#refugees-and-support-for-the-far-right-dinas-et.-al.-2018)):

- The setting is Greece during the 2015 refugee crisis, when islands closer to the Turkish coast experienced sudden and unexpected increases in refugee inflows during summer 2015 while other nearby islands saw much smaller inflows.
- Research question: what is the causal effect of refugee arrivals on support for the far-right Golden Dawn party?
- Outcome: municipal vote share for Golden Dawn (`gdvote`) in national elections.
- Identification approach: a DiD comparing treated vs. control municipalities before vs. after the crisis, with the paper focusing on the September 2015 election.
- Timing: in this file, all elections up to and including 2015 are treated as pre, and the 2016 election is treated as post.
- Data: 96 municipalities observed in 2012, 2013, 2015, and 2016; we compare exposed vs. non-exposed municipalities before and after 2015.
- Treatment coding in this file: `treatment` equals 1 only for treated municipalities in 2016 (post), and 0 otherwise; `ever_treated` is 1 for treated municipalities in all years.

Start by loading the data:

```{r}
load("dinas_golden_dawn.Rdata")
dinas_df <- as_tibble(muni)
```

Key variables:

- `treatment`: equals 1 if the observation is treated and in the post period (2016), 0 otherwise
- `ever_treated`: equals `TRUE` for treated municipalities in all periods, `FALSE` for controls
- `trarrprop`: per-capita refugee arrivals
- `gdvote`: Golden Dawn vote share
- `year`: election year (2012, 2013, 2015, 2016)

## Mapping variables to DiD notation

Which variable corresponds to $D_i$?
Which corresponds to $Y_{it}$?
Is this a panel or repeated cross-section?

### Answer

Start by exploring the variables in the data set. 

```{r}
names(dinas_df)
glimpse(dinas_df)
```

1. The variable that corresponds to $D_i$ is `ever_treated`, which is the indicator for belonging to the treatment group and is time-invariant (it equals 'TRUE' for treated municipalities in all years and 'FALSE' for control municipalities in all years). 

2. The variable that corresponds to $Y_{it}$ is `gdvote`, the outcome variable, which is the Golden Dawn vote share for a specific municipality in a specific year. 

3. This is a panel dataset, as the same units (in this case, the same 96 Greek municipalities) are observed repeatedly over multiple time points (2012, 2013, 2015, and 2016). The municipalities are fixed units, and `ever_treated` is constant within each municipality, so each municipality can be tracked over time. A repeated cross-section would instead involve different units sampled each year and no consistent unit identifier across time. 

As a check, each municipality should appear exactly four times (once for each election year).

```{r}
dinas_df |>
  group_by(municipality) |>
  summarize(count = n())
```

This is indeed the case, confirming a panel dataset. 

## Refugee inflows

Use `trarrprop` to document that treated municipalities experienced larger refugee inflows.

### Answer

Using refugee arrivals per capita, `trarrprop`, we will compare the average refugee arrivals between the two groups across the election years. 

```{r}
dinas_df |>
  group_by(year, ever_treated) |>
  summarize(avg_arrivals = mean(trarrprop, na.rm = TRUE))
```

In 2012, 2013, and 2015, the average refugee arrivals per capita (avg_arrivals) were effectively 0 for both groups. This suggests that there was no pre-existing refugee crisis that might have skewed political attitudes beforehand. In 2016, the control group (FALSE) remained at 0, while the treated group (TRUE) spiked to 2.85 mean arrivals per capita. Therefore, the treated municipalities experienced a sharp increase in refugee inflows in 2016, whereas the control municipalities and all pre-treatment periods saw zero arrivals, confirming the validity of the treatment classification.

## Naive post-period comparison

Using only 2016, compare Golden Dawn vote shares between treated and control municipalities.
Does this estimate have a causal interpretation?

### Answer

We will perform a simple comparison of the average vote shares for the treated and control municipalities in 2016.

```{r}
library(dplyr)

dinas_df |>
  filter(year == 2016) |>
  summarise(
    treated_mean  = mean(gdvote[ever_treated == TRUE], na.rm = TRUE),
    control_mean  = mean(gdvote[ever_treated == FALSE], na.rm = TRUE),
    difference    = treated_mean - control_mean
  )
```

In 2016, treated municipalities had Golden Dawn vote shares that were 2.74 percentage points higher than control municipalities (8.40% vs. 5.66%). However, this estimate does not have a causal interpretation because it does not account for pre-treatment differences between treated and control municipalities (treatment was not assigned at random). The observed gap may reflect baseline political differences rather than the effect of refugee inflows (potential selection bias).

## DiD (2015 vs 2016)

Compute the DiD estimate comparing 2015 (pre) and 2016 (post).
Do it manually and via regression.

### Answer

1. We will start by computing the relevant differences manually, using the mean function to construct the pre-treatment and post-treatment comparisons between treated and control groups.

```{r}
did_manual <- dinas_df |>
  filter(year %in% c(2015, 2016)) |>
  summarise(
    treat_2015 = mean(gdvote[ever_treated & year == 2015], na.rm = TRUE),
    treat_2016 = mean(gdvote[ever_treated & year == 2016], na.rm = TRUE),
    control_2015 = mean(gdvote[!ever_treated & year == 2015], na.rm = TRUE),
    control_2016 = mean(gdvote[!ever_treated & year == 2016], na.rm = TRUE)
  ) |>
  mutate(
    did = (treat_2016 - treat_2015) -
          (control_2016 - control_2015)
  )

did_manual
```

The DiD estimate is about 2.12 percentage points.

2. Next, we will compute the DiD estimate via regression. We start by creating a subset for the 2015/2016 comparison, a binary `post` indicator (1 if the year is 2016), then run the DiD on this subset using the lm() function. Recall that `treatment` is already the interaction (ever_treated * post).

```{r}
did_data <- dinas_df |> 
  filter(year %in% c(2015, 2016)) |>
  mutate(post = ifelse(year == 2016, 1, 0))

did_model <- lm(gdvote ~ ever_treated + post + treatment, data = did_data)

did_model |> 
  tbl_regression(
    intercept = TRUE,
    # This forces 2 decimal places for coefficients and CI
    estimate_fun = function(x) style_number(x, digits = 2),
    pvalue_fun = function(x) style_pvalue(x, digits = 2),
    label = list(
      ever_treated ~ "Treated Group (Di)",
      post ~ "Post-Period (Tt)",
      treatment ~ "DiD Estimate"
    )
  ) |>
  bold_labels()
```

The DiD estimate (coefficient for `treatment`) is 2.12, the same as computed above in the manual DiD. Therefore, using a DiD approach, we estimate that municipalities exposed to high refugee inflows experienced an additional 2.12 percentage point increase in Golden Dawn vote share in 2016 relative to control municipalities. This effect is statistically significant at the 5% level (p = 0.029).

## Plotting trends

Plot mean `gdvote` for treated and control municipalities across all years. What do you conclude about parallel trends?

### Answer

We will plot mean `gdvote` for both the treatment and control observations over time, along with the confidence intervals around each estimate. 

```{r}
library(ggplot2)

plot_data_ci <- dinas_df |>
  group_by(year, ever_treated) |>
  summarize(
    mean_gdvote = mean(gdvote, na.rm = TRUE),
    sd_gdvote = sd(gdvote, na.rm = TRUE),
    n = n(),
    .groups = 'drop'
  ) |>
  mutate(se = sd_gdvote / sqrt(n))

ggplot(plot_data_ci, aes(x = year, y = mean_gdvote, color = ever_treated, group = ever_treated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_gdvote - 1.96 * se, ymax = mean_gdvote + 1.96 * se), 
                width = 0.1) +
  geom_vline(xintercept = 2015.5, linetype = "dashed", color = "gray50") +
  scale_color_discrete(name = "Municipality Group", 
                       labels = c("FALSE" = "Control", "TRUE" = "Treatment")) +
  labs(title = "Golden Dawn Vote Share (95% CI)",
       subtitle = "Parallel Trends: Control vs. Treated Municipalities",
       x = "Year", y = "Average Vote Share (%)") +
  theme_minimal()
```

This plot provides strong visual support for the DiD identification strategy. Prior to the refugee crisis in 2016 (the three pre-treatment elections), treated and control municipalities evolve in roughly parallel trends in Golden Dawn support, with overlapping confidence intervals indicating no statistically significant differences in pre-treatment trajectories. Following the crisis, treated municipalities experienced a pronounced increase in vote share, diverging sharply from control municipalities. This suggests that the parallel trends assumption is reasonable, and the additional increase in Golden Dawn vote share in treated municipalities is plausibly due to the refugee crisis, not pre-existing differences. 

## Placebo DiD (2012 vs 2013)

Run a placebo DiD using 2012 and 2013. 
That means running a DiD using 2012 as pre and 2013 as post.
What do you find?
How do you interpret this result?

### Answer

We will run a placebo DiD test on 2012 vs. 2013 as another check to see if there were any false “treatment effects” before the refugee crisis, which would cast doubt on the parallel trends assumption.

```{r}
placebo_df <- dinas_df |>
  filter(year %in% c(2012, 2013)) |>
  mutate(post = ifelse(year == 2013, 1, 0))

placebo_did <- lm(gdvote ~ ever_treated * post, data = placebo_df)

tbl_placebo <- tbl_regression(
  placebo_did,
  intercept = TRUE,
  label = list(
    ever_treated = "Treated municipality",
    post = "Post period (2013)",
    `ever_treated:post` = "Placebo DiD (Treated × Post)"
  )
) 

tbl_placebo
```

The interaction term has an estimate of 0.0155 (~0.02), with a very large standard error and a p-value of 0.99, not statistically significant. There is no evidence of a differential change in Golden Dawn vote share between treated and control municipalities from 2012 to 2013. The near-zero estimate suggests treated and control municipalities followed parallel trends in this pre-treatment period.

This placebo test therefore supports the key parallel trends assumption in the main DiD analysis. Since no “effect” is found in the placebo period, it increases confidence that the observed treatment effect in 2016 reflects a true causal impact of refugee inflows, rather than pre-existing differences or divergent trends between treated and control municipalities.

## Two-way fixed effects

Estimate a TWFE regression using all years. Interpret the coefficient on `treatment`.

### Answer

We will estimate a TWFE model, which uses municipality fixed effects and year fixed effects. This controls for all permanent differences between municipalities (like geography or local culture) and all common shocks over time (like national economic shifts).

We will use the fixest package, a standard way to run fixed effects in R. 

```{r}
library(modelsummary)
library(fixest)

twfe_reg <- feols(gdvote ~ treatment | municipality + year, data = dinas_df)

modelsummary(
  twfe_reg,
  stars = TRUE,
  statistic = c("std.error", "p.value"),
  fmt = 3,
  estimate = "{estimate}"
)
```

The coefficient on treatment is 2.087, which is very close to our earlier 2-period DiD estimate of 2.12. This result suggests that exposure to the refugee inflows caused a 2.09 percentage point increase in Golden Dawn vote share (t = 5.31, p < 0.001). This confirms that the effect remains strong and significant when using the full panel of data and controlling for all fixed characteristics.

## Instrumental variables approach

The paper also uses a cross-sectional IV design. Key points from that setup:

- One observation per municipality (no time dimension in this file); 96 municipalities total.
- Outcome: change in Golden Dawn vote share between the January and September 2015 elections.
- Instrument: log distance to the Turkish coast, used to instrument for refugee inflows.
- First-stage intuition: municipalities closer to the Turkish coast saw larger refugee inflows.
- Variables in this file:
  - `treatment` (binary high exposure)
  - `trarrprop` (arrivals per capita)
  - `gdvote_change` (Jan–Sept 2015 change)
  - `logdist` (log distance to coast)

The data are in `dinas_golden_dawn_iv.Rdata`.

```{r}
load("dinas_golden_dawn_iv.Rdata")
iv_df <- as_tibble(muni)
```

## First stage

Estimate the first-stage relationship between the instrument and treatment.

### Answer

To estimate the first stage of this IV design, we need to regress our endogenous regressor (treatment) on our instrument (logdist).

```{r}
first_stage <- lm(treatment ~ logdist, data = iv_df)

first_stage |> 
  tbl_regression(
    label = list(logdist ~ "Log Distance to Turkish Coast"),
    estimate_fun = function(x) style_number(x, digits = 2)
  ) |>
  add_glance_source_note(label = list(statistic = "F-statistic")) |>
  bold_labels()
```

A one-unit increase in log distance from the Turkish coast reduces refugee arrivals per capita by about 0.21 units (p-value < 0.001), with an F-statistic value of about 137, far above the conventional threshold of 10. Therefore, municipalities farther from the Turkish coast received significantly fewer refugees, this matches the identification logic of the paper. 

## 2SLS

Use `iv_robust` to estimate the causal effect of refugee exposure on vote share change. Do this for both the binary and continuous treatment measures. Interpret the coefficients and compare them to the DiD estimates above.

### Answer

To estimate the IV model with robust standard errors, we use the iv_robust function from the estimatr package. This performs Two-Stage Least Squares (2SLS), using logdist to extract the exogenous variation in refugee exposure. We will do this for both the binary `treatment` variable and the continuous `trarrprop` variable. 

Starting with the binary treatment regression.

```{r}
iv_binary <- iv_robust(
  gdvote_change ~ treatment | logdist,
  data = iv_df
)

cm = c("treatment" = "Refugee Treatment (Binary)", "(Intercept)" = "Constant")

msummary(iv_binary,
         stars = TRUE,
         fmt = 2,
         coef_map = cm,
         gof_omit = 'AIC|BIC|Log.Lik|F|RMSE',
         title = "Table: Causal Effect of Refugee Exposure (IV Design)")
```

Municipalities with high refugee exposure experienced a 2.08 percentage point larger increase in Golden Dawn vote share (Jan–Sept 2015), causally attributable to refugee inflows. This is statistically significant at the 1% level.

Next, continuous treatment regression.

```{r}
iv_continuous <- iv_robust(
  gdvote_change ~ trarrprop | logdist,
  data = iv_df
)

cm_cont = c("trarrprop" = "Arrivals per Capita", "(Intercept)" = "Constant")

msummary(iv_continuous,
         stars = TRUE,
         fmt = 2,
         coef_map = cm_cont,
         gof_omit = 'AIC|BIC|Log.Lik|F|RMSE',
         title = "Table: Causal Effect of Refugee Arrival Intensity (IV Design)")
```

A one-unit increase in refugee arrivals per capita increased Golden Dawn vote share by approximately 0.74 percentage points (statistically significant at the 1% level). This gives a marginal causal effect rather than a treatment-group contrast.

The IV identifies a Local Average Treatment Effect (LATE), therefore these estimates reflect the causal effect for municipalities whose refugee exposure was influenced by their distance to the Turkish coast, rather than necessarily the effect for all municipalities. Notably, the binary IV estimate at 2.08 is nearly identical to the DiD estimate at 2.09 (the LATE and ATT are numerically very close), providing strong and consistent evidence that refugee inflows causally increased support for the far-right Golden Dawn party. The similarity between the IV and DiD results suggests limited bias from selection, supports the plausibility of the parallel trends assumption in the panel analysis, and reinforces the credibility of distance to the coast as a valid geographic instrument.

## IV assumptions

1. State the exclusion restriction for `logdist`.
2. Is it plausible in this setting?

### Answer

1. For log distance to the Turkish coast (logdist) to be a valid instrument, it must satisfy the following: the distance between a Greek municipality and the Turkish coast must affect the change in Golden Dawn’s vote share only through its impact on refugee arrivals, and not through any other direct or indirect channel.

2. This assumption is plausible because refugee arrivals were geographically driven and largely determined by proximity to Turkey (the "assignment" of refugees to specific islands was a matter of proximity, not local political leanings). Additionally, the strong first stage confirms that distance predicts refugee exposure, it is difficult to think of another factor that is perfectly correlated with distance to Turkey and changed so drastically in those exact 9 months to produce a 2-point swing in voting. However, the restriction could be violated if municipalities closer to Turkey experienced other concurrent shocks that independently affected far-right support (for example, border/island municipalities may differ economically (tourism, trade, migration history) and proximity to Turkey might influence national identity or security attitudes even absent refugee inflows). Therefore, while plausible, the assumption is not directly testable and must be justified substantively.



